Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-04-20 doc/thesis/intro.tex
Few more references for phylodynamics

Due to the relationship between the aforementioned processes, there is a degree
of correspondence between viral phylogenies and transmission
trees~\autocite{ypma2013relating, kenah2015algorithms, kenah2016molecular}. In
particular, the transmission process is quite similar to \defn{allopatric
speciation}~\autocite{coyne2004speciation}, where genetic divergence follows
the geographic isolation of a sub-population of organisms. Thus, transmission,
which is represented as branching in the transmission tree, causes branching in
the viral phylogeny as well. Similarly, the removal of an individual from the
transmission tree causes the extinction of their viral lineage in the
phylogeny. Consequently, the topology of the viral phylogeny is sometimes used
as a proxy for the topology of the transmission tree. Modern likelihood-based
methods of phylogenetic reconstruction~\autocite[\eg][]{price2010fasttree,
stamatakis2014raxml} produce unrooted trees whose branch lengths measure
genetic distance in units of expected substitutions per site. On the other
hand, transmission trees are rooted, and have branches measuring calendar
time~\autocite{pybus2009evolutionary}. Therefore, estimating a transmission
tree from a viral phylogeny requires the phylogeny to be rooted and
time-scaled. Methods for performing this process include root-to-tip
regression~\autocite{shankarappa1999consistent, korber2000timing,
drummond2003inference}, which we apply in this work, and least-square
dating~\autocite{to2015fast}. Alternatively, the tree may be rooted separately
with an outgroup~\autocite{li1988rates} before time-scaling.

A caveat of estimating transmission trees in this manner is that the
correspondence between the topologies of the viral phylogeny and transmission
tree is not necessarily exact~\autocite{ypma2013relating}. Due to intra-host
diversity, the viral strain which is transmitted may have split from another
lineage within the donor long before the transmission event occurred. Hence,
the branching point in the viral phylogeny may be much earlier than that in the
transmission tree. Another possibility is that one host transmitted to two or
more recipients, but the lineages they each received originated within the
donor host in a different order than that in which the transmissions occurred.
In this case, the topology of the transmission tree and the viral phylogeny
will be mismatched. In practice, this discordance has not proven an
insurmountable problem: for example, \textcite{leitner1996accurate} were able
to accurately recover a known transmission tree using a viral phylogeny. The
problem of accurately estimating transmission trees is an ongoing area of
research~\autocite{cottam2008integrating, jombart2011reconstructing,
ypma2012unravelling, didelot2014bayesian}.

>>>>

Due to the relationship between the aforementioned processes, there is a degree
of correspondence between viral phylogenies and transmission
trees~\autocite{ypma2013relating, kenah2015algorithms, kenah2016molecular}. In
particular, the transmission process is quite similar to \defn{allopatric
speciation}~\autocite{coyne2004speciation}, where genetic divergence follows
the geographic isolation of a sub-population of organisms. Thus, transmission,
which is represented as branching in the transmission tree, causes branching in
the viral phylogeny as well. Similarly, the removal of an individual from the
transmission tree causes the extinction of their viral lineage in the
phylogeny. Consequently, the topology of the viral phylogeny is sometimes used
as a proxy for the topology of the transmission
tree~\autocite{hall2015epidemic}. Modern likelihood-based methods of
phylogenetic reconstruction~\autocite[\eg][]{price2010fasttree,
stamatakis2014raxml} produce unrooted trees whose branch lengths measure
genetic distance in units of expected substitutions per site. On the other
hand, transmission trees are rooted, and have branches measuring calendar
time~\autocite{pybus2009evolutionary}. Therefore, estimating a transmission
tree from a viral phylogeny requires the phylogeny to be rooted and
time-scaled. Methods for performing this process include root-to-tip
regression~\autocite{shankarappa1999consistent, korber2000timing,
drummond2003inference}, which we apply in this work, and least-square
dating~\autocite{to2015fast}. Alternatively, the tree may be rooted separately
with an outgroup~\autocite{li1988rates} before time-scaling.

A caveat of estimating transmission trees in this manner is that the
correspondence between the topologies of the viral phylogeny and transmission
tree is not necessarily exact~\autocite{ypma2013relating, romero2014timing}.
Due to intra-host diversity, the viral strain which is transmitted may have
split from another lineage within the donor long before the transmission event
occurred. Hence, the branching point in the viral phylogeny may be much earlier
than that in the transmission tree. Another possibility is that one host
transmitted to two or more recipients, but the lineages they each received
originated within the donor host in a different order than that in which the
transmissions occurred.  In this case, the topology of the transmission tree
and the viral phylogeny will be mismatched. In practice, this discordance has
not proven an insurmountable problem: for example,
\textcite{leitner1996accurate, paraskevis2004phylogenetic} were able to
accurately recover a known transmission tree using a viral phylogeny. The
problem of accurately estimating transmission trees is an ongoing area of
research~\autocite{cottam2008integrating, jombart2011reconstructing,
ypma2012unravelling, didelot2014bayesian}.
